---
title: "DNMTs: Dimensionality reduction using Seurat"
---
  
```{r load_modules, echo=FALSE, include=FALSE}
library(Seurat)
```

# Define settings

Define I/O
```{r define_io, echo=FALSE}
source("/Users/ricard/10x_gastrulation_DNMTs/settings.R")
io$outdir <- paste0(io$basedir,"/results/dimensionality_reduction")
```

Define options
```{r}
opts$batches <- c(
  # "SIGAA6_E85_2_Dnmt3aKO_Dnmt3b_WT_L001", 
  # "SIGAB6_E85_3_Dnmt3aWT_Dnmt3b_WT_L002", 
  # "SIGAC6_E85_5_Dnmt3aKO_Dnmt3b_Het_L003", 
  # "SIGAD6_E85_8_Dnmt3aHet_Dnmt3b_KO_L004",
  "15_E8_5_D3A_WT_D3B_WT_L007"
  # "17_E8_5_D3A_KO_D3B_WT_L008"
  # "2_E8_5_D3A_WT_D3B_KO_L003"
  # "3_E8_5_D3A_HET_D3B_WT_L004"
  # "7_E8_5_D3A_WT_D3B_KO_L005"
  # "8_E8_5_D3A_KO_D3B_KO_L006"
)
```

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- fread(io$metadata) %>%
  .[pass_QC==TRUE & batch%in%opts$batches]
table(sample_metadata$batch)
```

# Load data

Load RNA expression data as Seurat object
```{r load_data, echo=FALSE}
seurat <- readRDS(io$seurat)[,sample_metadata$cell]
dim(seurat)
```

Add metadata to the Seurat object
```{r}
foo <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(seurat),]

stopifnot(colnames(seurat) == rownames(foo))
# seurat <- AddMetaData(seurat, foo)  # not working????
seurat@meta.data <- foo
```


# Parse data 

Remove genes that are not expressed
```{r}
foo <- Matrix::rowMeans(seurat@assays$RNA@counts)
seurat <- subset(seurat, features = names(which(foo>1e-6)))
```

Normalise
```{r message=FALSE}
# SCTransform
seurat <- SCTransform(seurat, vars.to.regress = "percent.mt")

# LogNormalize
# seurat <- NormalizeData(seurat, normalization.method = "LogNormalize")

# Scale
```

Scale (only if using logNorm)
```{r}
# seurat <- ScaleData(seurat, vars.to.regress = "nFeature_RNA")
```

Change gene names from ENSEMBL to symbols
```{r}
# gene_metadata <- fread(io$gene_metadata) %>% .[,c("ens_id","symbol")] %>%
#   .[symbol!="" & ens_id%in%rownames(seurat)]
# 
# seurat <- seurat[rownames(seurat)%in%gene_metadata$ens_id,]
# foo <- gene_metadata$symbol; names(foo) <- gene_metadata$ens_id
# new.names <- foo[rownames(seurat)]
# 
# # Sanity cehcks
# stopifnot(sum(is.na(new.names))==0)
# stopifnot(sum(duplicated(new.names))==0)
# 
# new.names[duplicated(new.names)]
# 
# # Rename seurat objects manually
# rownames(seurat@assays$SCT@counts) <- new.names
# rownames(seurat@assays$SCT@data) <- new.names
# # rownames(seurat@assays$SCT@scale.data) <- new.names
# seurat[["SCT"]]@meta.features <- data.frame(row.names=new.names)
```


# Select HVG

Option 1, using FindVariableFeatures
```{r}
seurat <- FindVariableFeatures(seurat, selection.method = 'vst', nfeatures = 1500, assay = "SCT")
# length(seurat@assays$RNA@var.features)
```

Option 2, using precomputed marker genes
```{r}
# marker_genes.dt <- fread(io$atlas.marker_genes)
# hvg <- unique(marker_genes.dt$ens_id)
# hvg <- hvg[hvg%in%rownames(seurat)]
# 
# seurat@assays$RNA@var.features <- hvg
```

# Dimensionality reduction

PCA 
```{r message=FALSE}
seurat <- RunPCA(seurat, 
  features = VariableFeatures(seurat, assay="SCT"),
  assay = "SCT",
  npcs = 25,
  verbose = FALSE
)
```

```{r}
cor(seurat$percent.mt,seurat@reductions$pca@cell.embeddings)
```

<!-- Correlation between PCs and batch -->
```{r}
# r <- cor(
#   seurat@reductions$pca@cell.embeddings,
#   as.numeric(as.factor(seurat@meta.data$batch))
# )[,1]
# r
```

UMAP
```{r}
seurat <- RunUMAP(seurat, 
  dims = 1:25,
  reduction = "pca",
  n.neighbors = 30,
  min.dist = 0.3
)
```

# Plot 

```{r}
Idents(seurat) <- "celltype.mapped"
# Idents(seurat) <- "celltype.mapped.iterative"

p <- DimPlot(seurat, label = FALSE, reduction = 'pca', pt.size=2) + 
# p <- DimPlot(seurat, label = FALSE, reduction = 'pca', cols = opts$celltype.colors, pt.size=2) + 
  # NoLegend() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

print(p)
```

```{r}
Idents(seurat) <- "celltype.mapped"

p <- DimPlot(seurat, label = FALSE, reduction = 'pca') + 
  NoLegend() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

print(p)
```

Explore marker genes
```{r}
# genes.to.plot <- c("Apom")
# genes.to.plot <- c("Hbb-y","Hba-a2","Mgst3")
genes.to.plot <- c("Hbb-y")
FeaturePlot(seurat, features = genes.to.plot, reduction='pca', pt.size=2)

```

```{r}
VlnPlot(seurat, features=genes.to.plot, group.by = "celltype.mapped") + NoLegend()
```



