---
title: "Dimensionality reduction"
---
  
```{r load_modules, echo=FALSE, include=FALSE}
library(Seurat)
```

# Define settings

Define I/O
```{r define_io, echo=FALSE}
source("/Users/ricard/10x_gastrulation_DNMTs/settings.R")
io$outdir <- paste0(io$basedir,"/results/dimensionality_reduction")
```

Define options
```{r}
opts$batches <- c(
  # "E125_DNMT3A_HET_A_L001", 
  # "E125_DNMT3A_HET_A_L003", 
  # "E125_DNMT3A_KO_B_L002", 
  # "E125_DNMT3A_KO_E_L004", 
  "SIGAA6_E85_2_Dnmt3aKO_Dnmt3b_WT_L001",
  "SIGAB6_E85_3_Dnmt3aWT_Dnmt3b_WT_L002",
  "SIGAC6_E85_5_Dnmt3aKO_Dnmt3b_Het_L003",
  "SIGAD6_E85_8_Dnmt3aHet_Dnmt3b_KO_L004"
)

opts$celltypes = c(
	# "Epiblast",
	# "Primitive_Streak",
	# "Caudal_epiblast",
	# "PGC",
	# "Anterior_Primitive_Streak",
	# "Notochord",
	# "Def._endoderm",
	# "Gut",
	# "Nascent_mesoderm",
	# "Mixed_mesoderm",
	# "Intermediate_mesoderm",
	# "Caudal_Mesoderm",
	# "Paraxial_mesoderm",
	# "Somitic_mesoderm",
	# "Pharyngeal_mesoderm",
	# "Cardiomyocytes",
	# "Allantois",
	# "ExE_mesoderm",
	# "Mesenchyme",
	"Haematoendothelial_progenitors",
	"Endothelium",
	"Blood_progenitors_1",
	"Blood_progenitors_2",
	"Erythroid1",
	"Erythroid2",
	"Erythroid3"
	# "NMP",
	# "Rostral_neurectoderm",
	# "Caudal_neurectoderm",
	# "Neural_crest",
	# "Forebrain_Midbrain_Hindbrain",
	# "Spinal_cord",
	# "Surface_ectoderm",
	# "Visceral_endoderm",
	# "ExE_endoderm",
	# "ExE_ectoderm",
	# "Parietal_endoderm"
)
```

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- sample_metadata %>%
  .[batch%in%opts$batches & celltype.mapped%in%opts$celltypes]
table(sample_metadata$batch)
```

# Load data

Load RNA expression data as Seurat object
```{r load_data, echo=FALSE}
seurat <- readRDS(io$seurat)[,sample_metadata$cell]
dim(seurat)
```

Add metadata to the Seurat object
```{r}
foo <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(seurat),]

stopifnot(colnames(seurat) == rownames(foo))
# seurat <- AddMetaData(seurat, foo)  # not working????
seurat@meta.data <- foo
```


# Parse data 

Remove genes that are not expressed
```{r}
foo <- Matrix::rowMeans(seurat@assays$RNA@counts)
seurat <- subset(seurat, features = names(which(foo>1e-6)))
```

Normalise
```{r}
# SCTransform
# seurat <- SCTransform(seurat)

# LogNormalize
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize")

# Scale
seurat <- ScaleData(seurat)
```

Change gene names from ENSEMBL to symbols
```{r}
gene_metadata <- fread(io$gene_metadata) %>% .[,c("ens_id","symbol")] %>%
  .[symbol!="" & ens_id%in%rownames(seurat)]

seurat <- seurat[rownames(seurat)%in%gene_metadata$ens_id,]
foo <- gene_metadata$symbol; names(foo) <- gene_metadata$ens_id
new.names <- foo[rownames(seurat)]

# Sanity cehcks
stopifnot(sum(is.na(new.names))==0)
stopifnot(sum(duplicated(new.names))==0)

new.names[duplicated(new.names)]

# Rename seurat objects manually
rownames(seurat@assays$RNA@counts) <- new.names
rownames(seurat@assays$RNA@data) <- new.names
rownames(seurat@assays$RNA@scale.data) <- new.names
seurat[["RNA"]]@meta.features <- data.frame(row.names=new.names)
```


# Select HVG

Option 1, using FindVariableFeatures
```{r}
seurat <- FindVariableFeatures(seurat, selection.method = 'vst', nfeatures = 1000)
# length(seurat@assays$RNA@var.features)
```

Option 2, using precomputed marker genes
```{r}
# marker_genes.dt <- fread(io$atlas.marker_genes)
# hvg <- unique(marker_genes.dt$ens_id)
# hvg <- hvg[hvg%in%rownames(seurat)]
# 
# seurat@assays$RNA@var.features <- hvg
```

# Dimensionality reduction

PCA 
```{r message=FALSE}
seurat <- RunPCA(seurat, 
  features = VariableFeatures(seurat),
  npcs = 25,
  verbose = FALSE
)
```

<!-- Correlation between PCs and batch -->
```{r}
# r <- cor(
#   seurat@reductions$pca@cell.embeddings,
#   as.numeric(as.factor(seurat@meta.data$batch))
# )[,1]
# r
```

UMAP
```{r}
seurat <- RunUMAP(seurat, 
  dims = 1:25,
  # dims = which(r<=0.5),
  reduction = "pca"
)
```

# Plot 

```{r}
Idents(seurat) <- "celltype.mapped"
DimPlot(seurat, label = FALSE, reduction = 'umap', cols = opts$celltype.colors)# + NoLegend()
```

```{r}
Idents(seurat) <- "class"
DimPlot(seurat, label = FALSE, reduction = 'umap')# + NoLegend()
```

Explore marker genes
```{r}
# genes.to.plot <- c("Apom")
# FeaturePlot(seurat, features = genes.to.plot, reduction='umap')
```



