---
title: "Sex determination using RNA expression counts of genes on the Y chromosome."
---

```{r echo=FALSE, include=FALSE}
library(scater)
library(ggpubr)
```

```{r define_opts, echo=FALSE, include=FALSE}
if (grepl("ricard",Sys.info()['nodename'])) {
  source("/Users/ricard/10x_gastrulation_DNMTs/settings.R")
} else {
  stop("Computer not recognised")
}
io$outdir <- paste0(io$basedir,"/results/sex")
```

Define options
```{r}
opts$batches <- c(
  "SIGAA6_E85_2_Dnmt3aKO_Dnmt3b_WT_L001", 
  "SIGAB6_E85_3_Dnmt3aWT_Dnmt3b_WT_L002", 
  "SIGAC6_E85_5_Dnmt3aKO_Dnmt3b_Het_L003", 
  "SIGAD6_E85_8_Dnmt3aHet_Dnmt3b_KO_L004",
  "15_E8_5_D3A_WT_D3B_WT_L007",
  "17_E8_5_D3A_KO_D3B_WT_L008",
  "2_E8_5_D3A_WT_D3B_KO_L003",
  "3_E8_5_D3A_HET_D3B_WT_L004",
  "7_E8_5_D3A_WT_D3B_KO_L005",
  "8_E8_5_D3A_KO_D3B_KO_L006"
  
  # "E125_DNMT3A_HET_A_L001", 
  # "E125_DNMT3A_HET_A_L003", 
  # "E125_DNMT3A_KO_B_L002", 
  # "E125_DNMT3A_KO_E_L004", 
  # "A_E12_5_D3a_Het_L001",
  # "B_E12_5_D3a_KO_L002"
)
```

Update sample metadata
```{r}
sample_metadata <- sample_metadata[batch%in%opts$batches]
dim(sample_metadata)
```


# Load data

Load SingleCellExperiment
```{r load_expr, echo=FALSE, include=FALSE}
sce <- readRDS(io$sce)[,sample_metadata$cell]
dim(sce)
```

Load gene metadata
```{r}
gene_metadata <- fread(io$gene_metadata) %>% 
  .[ens_id%in%rownames(sce)]
```

# Parse data

Group genes by chromosome
```{r}
genes.chrY <- gene_metadata[chr=="chrY",ens_id]
genes.chrX <- gene_metadata[chr=="chrX",ens_id]
genes.chr10 <- gene_metadata[chr=="chr10",ens_id]
```

Manual filtering
```{r}
# For some reason Erdr1 is predicted as Ychr, but the last version of ENSEMBL is in the Xchr
genes.chrY <- genes.chrY[!genes.chrY=="ENSMUSG00000096768"]
```


Create data.table from SingleCellExperiment object
```{r}
# dt <- counts(sce[c(genes.chrY,genes.chrX,genes.chr10)]) %>% as.matrix %>% t %>%
#   as.data.table(keep.rownames="cell") %>% 
#   melt(id.vars="cell", value.name="counts", variable.name="ens_id") %>%
#   merge(sample_metadata[,c("cell","batch")], by="cell")#  %>%
#   # merge(gene_metadata[,c("chr","ens_id","symbol")], by="ens_id")
```

```{r}
dt <- opts$batches %>% map(function(i) {
  sce.filt <- sce[,sce$batch==i] %>% .[c(genes.chrY,genes.chrX,genes.chr10),]
  dt <- data.table(
    ens_id = rownames(sce.filt),
    counts = counts(sce.filt) %>% Matrix::rowSums()
  ) %>% .[,batch:=i]
}) %>% rbindlist %>% merge(gene_metadata[,c("chr","ens_id","symbol")], by="ens_id")
```

# Plot

## Individual genes chrY

```{r}
to.plot <- dt[chr=="chrY"] %>% 
  # .[,.(counts=sum(counts)),by=c("batch","chr","ens_id","symbol")] %>%
  .[,.(counts=sum(counts)),by=c("batch","chr","ens_id")] %>%
  .[,mean:=mean(counts),by="ens_id"] %>% .[mean>0] %>% .[,mean:=NULL]
```


```{r}
# p <- ggbarplot(to.plot, x="symbol", y="counts", facet="batch") +
p <- ggbarplot(to.plot, x="ens_id", y="counts", facet="batch") +
  labs(x="", y="Read counts") +
  theme(
  axis.text.x = element_text(colour="black",size=rel(0.7), angle=45, hjust=1, vjust=1),
  axis.text.y = element_text(colour="black",size=rel(0.8))
  )

# pdf(sprintf("%s/pdf/sex_ychr_expr_per_gene.pdf",io$outdir), width=8, height=6, useDingbats = F)
print(p)
# dev.off()
```

## Agregated genes

```{r}
to.plot <- dt %>% 
  .[,.(counts=sum(counts)),by=c("batch","chr")] %>%
  dcast(batch~chr, value.var="counts") %>%
  .[,ratioY:=chrY/chr10] %>% .[,ratioX:=chrX/chr10]
```

Sex assignment
```{r}
opts$threshold.ratioY <- 1e-3 
to.plot %>% .[,sex:=c("female","male")[as.numeric(ratioY>=opts$threshold.ratioY)+1]]
```

Save
```{r}
fwrite(to.plot, paste0(io$outdir,"/sex_assignment.txt.gz"))
```


Scatter of chrY/chr1 vs chrX/chr1 count ratio per embryo
```{r}
# ggscatter(to.plot, x="ratioX", y="ratioY") +
#   labs(x="chrX/chr1 ratio", y="chrY/chr1 ratio") +
#   theme(
#     # axis.text.x = element_text(colour="black",size=rel(0.8), angle=45, hjust=1, vjust=1),
#     # axis.title.x = element_blank(),
#     # axis.ticks.x = element_blank()
#   )
```

Barplots of chrXY/chr1 count ratio per embryo
```{r}
p <- ggbarplot(to.plot, x="batch", y="ratioY", fill="sex", sort.val = "asc", palette="Dark2") +
  labs(x="", y="chrY/chr1 counts ratio") +
  theme(
    legend.position = "right",
    # axis.text.x = element_text(colour="black",size=rel(0.8), angle=45, hjust=1, vjust=1),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# pdf(sprintf("%s/pdf/sex_ychr_expr_aggregated.pdf",io$outdir), width=8, height=6, useDingbats = F)
print(p)
# dev.off()
```

# Plot expression of Xist

```{r}
to.plot <- dt[ens_id%in%"ENSMUSG00000086503"]

p <- ggbarplot(to.plot, x="batch", y="counts", fill="gray70") +
  labs(x="", y="Xist expression") +
  theme(
    axis.text.x = element_text(colour="black",size=rel(0.8), angle=45, hjust=1, vjust=1),
  )

# pdf(sprintf("%s/pdf/Xist_expr.pdf",io$outdir), width=8, height=6, useDingbats = F)
print(p)
# dev.off()
```

